Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-01-28 Master Thesis/synopsis.tex
Replace max unramified with absolute inertia

\begin{theorem}\label{theorem-maxunramified}
Let $K$ be a henselian field with respect to $v: K\to\Gamma\cup\{\infty\}$. Let $K^\text{nr}$ be the maximal unramified extension of $K$ with the unique extension $w$ of $v$. Then:
\begin{enumerate}[(i)]
\item $w$ has value group $\Gamma$ and residue class field $k(v)^\text{s}$.
\item $K^\text{nr}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{nr}/K)\cong G_{k(v)}$. There exists an inclusion-preserving one-to-one correspondence: 
\begin{align*}
\{ \text{Intermediate extensions in $K^\text{nr}/K$} \} &\longrightarrow \{\text{Separable extensions of $k(v)$} \}\\ 
L &\longmapsto k(v_L)  
\end{align*}
where $v_L$ is the unique extension of $v$ to $L$.
\end{enumerate}
\end{theorem}

@@

Furthermore, if $E^{\text{nr}}$ and $E'^{\text{nr}}$ denote the maximal unramified field extensions of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{nr}/E) \to\operatorname{Gal}(E'^{\text{nr}}/E')\cong G_{k(v')} \]
which is induced by a unique isomorphism $\varphi_v^\text{s}:k(v')^\text{s}\to k(v)^\text{s}$. In particular $\varphi_v=\varphi_v^\text{s}|_{k(v')}:k(v')\to k(v)$ is an isomorphism.
\end{lemma}

@@

Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. By Lemma \ref{2.2} the following statements are equivalent:
\begin{enumerate}[(i)]
\item The composite $K_{w_1}K_{w_2}$ is not algebraically closed.
\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v(w_1) = v(w_2)$.
\end{enumerate}
In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$. By Theorem \ref{theorem-maxunramified} $K_{w_1}K_{w_2}$ is the maximal unramified field extension $K_v^\text{nr}$ of $K_v$. Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:
\[ v(w_1)=v(w_2) \iff v(\sigma(w_1)) = v(\sigma(w_2)) \]
This shows that $\sigma$ also induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$ and for all $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ we have $\sigma(v(w)) = v(\sigma(w))$.

Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$. Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$. This shows that $K'_{v'} = (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$. Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
\[ \sigma(G_{K^\text{nr}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{nr}_{v'}} \]
Therefore, $\sigma$ induces with Theorem \ref{theorem-maxunramified} the following isomorphism:
\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{nr}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{nr}}\cong G_{k(v')} \]
The rest follows from the Theorem of Neukirch-Uchida.
\end{proof}

@@

\begin{proof}
\begin{enumerate}[(i)]
\item Consider the maximal unramified extension $L = K_P^\text{nr}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$. Now $k(w) = k(v_P)^\text{s}$. Since $w$ is unramified over $v$, $w$ is discrete as well. We have the following exact sequence:
\[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$. Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$. Now the following commutative diagram with exact columns gives us the result:
\[ \begin{tikzcd}
1 \ar[d] & 1 \ar[d]\\
U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\
K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
L'^\times /n \ar[r, "\varphi_n"'] &  L^\times /n
\end{tikzcd}\]
Note that $(K^\text{nr}_P)' = K'^{\text{nr}}_{P'}$ as proven in Lemma \ref{3.2}.
\item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
\[\begin{tikzcd}
k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[rr, hook] && K_P^\times/n
\end{tikzcd}\]
coincides with the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$. The same holds for $P'$ due to Lemma \ref{3.2}. Therefore, the assertion follows from the commutativity of the following diagram:
\[ \begin{tikzcd}
k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
\end{tikzcd}  \qedhere\]
\end{enumerate}
\end{proof}

>>>>

\begin{theorem}\label{theorem-absoluteinertia}
Let $K$ be a henselian field with respect to $v: K\to\Gamma\cup\{\infty\}$. Let $K^\text{t}$ be the absolute inertia field of $K$ with the unique extension $w$ of $v$. Then:
\begin{enumerate}[(i)]
\item $w$ has value group $\Gamma$ and residue class field $k(v)^\text{s}$.
\item $K^\text{t}$ is a Galois extension of $K$ and $\operatorname{Gal}(K^\text{t}/K)\cong G_{k(v)}$. There exists an inclusion-preserving one-to-one correspondence: 
\begin{align*}
\{ \text{Intermediate extensions in $K^\text{t}/K$} \} &\longrightarrow \{\text{Separable extensions of $k(v)$} \}\\ 
L &\longmapsto k(v_L)  
\end{align*}
where $v_L$ is the unique extension of $v$ to $L$.
\end{enumerate}
\end{theorem}

@@

Furthermore, if $E^{\text{t}}$ and $E'^{\text{t}}$ denote the absolute inertia field of $E$ and $E'$ respectively, then $\sigma$ induces an isomorphism:
\[ G_{k(v)}\cong\operatorname{Gal}(E^\text{t}/E) \to\operatorname{Gal}(E'^{\text{t}}/E')\cong G_{k(v')} \]
which is induced by a unique isomorphism $\varphi_v^\text{s}:k(v')^\text{s}\to k(v)^\text{s}$. In particular $\varphi_v=\varphi_v^\text{s}|_{k(v')}:k(v')\to k(v)$ is an isomorphism.
\end{lemma}

@@

Let $w_1,w_2 \in\operatorname{Spv}_\text{d}^{(2)}(K)$ be two different valuations. By Lemma \ref{2.2} the following statements are equivalent:
\begin{enumerate}[(i)]
\item The composite $K_{w_1}K_{w_2}$ is not algebraically closed.
\item $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ and $v(w_1) = v(w_2)$.
\end{enumerate}
In this case, $k(v')$ is separably closed where $v'$ is the unique extension of the valuation on $K_v$ to $K_{w_1}K_{w_2}$ where $v=v(w_1)$. By Theorem \ref{theorem-absoluteinertia} $K_{w_1}K_{w_2}$ is the absolute inertia field $K_v^\text{t}$ of $K_v$. Furthermore, $K'_{\sigma(w_1)}K'_{\sigma(w_2)} = (K_{w_1}K_{w_2})' \neq \overline{K'}$ and again by Lemma \ref{2.2} we get: \[\sigma(w_1),\sigma(w_2)\in\operatorname{Spv}_\text{rd}^{(2)}(K)\quad \text{and} \quad v(\sigma(w_1)) = v(\sigma(w_2))\]
Thus $\sigma: \operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ maps $\operatorname{Spv}_\text{rd}^{(2)}(K)$ onto $\operatorname{Spv}_\text{rd}^{(2)}(K')$ such that:
\[ v(w_1)=v(w_2) \iff v(\sigma(w_1)) = v(\sigma(w_2)) \]
This shows that $\sigma$ also induces a bijection $\sigma:\operatorname{Spv}_\text{d}^{(1)}(K)\to\operatorname{Spv}_\text{d}^{(1)}(K')$ and for all $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ we have $\sigma(v(w)) = v(\sigma(w))$.

Let $v\in\operatorname{Spv}_\text{d}^{(1)}(K)$ and let $N$ be the normal subgroup of $G_{K_v}$ generated by all subgroups $G_{K_w}$ for $w\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $v(w) = v$. Since a global field has no non-trivial algebraic extension in which every prime splits completely, we get $N=G_{K_v}$. This shows that $K'_{v'} = (K_v)'$ is a henselization of $K'$ with respect to $v'=\sigma(v)$. Furthermore, let $w_1,w_2\in\operatorname{Spv}_\text{rd}^{(2)}(K)$ with $w_1\neq w_2$ and $v=v(w_1)=v(w_2)$, then:
\[ \sigma(G_{K^\text{t}_v}) = \sigma(G_{K_{w_1}}\cap G_{K_{w_2}}) = G_{K'_{w'_1}}\cap G_{K'_{w'_2}} = G_{(K')^\text{t}_{v'}} \]
Therefore, $\sigma$ induces with Theorem \ref{theorem-absoluteinertia} the following isomorphism:
\[ G_{k(v)} \cong G_{K_v}/G_{K_v^\text{t}} \to G_{K'_{v'}}/G_{(K')_{v'}^\text{t}}\cong G_{k(v')} \]
The rest follows from the Theorem of Neukirch-Uchida.
\end{proof}

@@

\begin{proof}
\begin{enumerate}[(i)]
\item Consider the absolute inertia field $L = K_P^\text{t}$ of $K_P$ and let $w$ be the unique extension of $v_P$ in $L$. Now $k(w) = k(v_P)^\text{s}$. Since $w$ is unramified over $v$, $w$ is discrete as well. We have the following exact sequence:
\[ 1 \longrightarrow U_w/n \longrightarrow L^\times /n\longrightarrow \mathbb{Z}/n\mathbb{Z}\longrightarrow 0 \]
Since $k(w)$ is separably closed, with $\operatorname{char} k(w)=0$ and Hensel's lemma we can see that $U_w$ is a divisible group, hence we get an isomorphism $L^\times /n\cong \mathbb{Z}/n\mathbb{Z}$ induced by $w$. Therefore, the canonical map $K^\times/n\to L^\times/n$ can be identified with the map $K^\times/n \to\mathbb{Z}/n\mathbb{Z}$ induced by $v_P$. Now the following commutative diagram with exact columns gives us the result:
\[ \begin{tikzcd}
1 \ar[d] & 1 \ar[d]\\
U_{P'}/n \ar[r, dashed]\ar[d] & U_P/n\ar[d]\\
K'^\times /n \ar[r, "\varphi_n"]\ar[d] & K^\times/n\ar[d]\\
L'^\times /n \ar[r, "\varphi_n"'] &  L^\times /n
\end{tikzcd}\]
Note that $(K^\text{t}_P)' = K'^{\text{t}}_{P'}$ as proven in Lemma \ref{3.2}.
\item Consider the field extension $K_P/K$. Without loss of generality, we can assume $K=K_P$. Since $\operatorname{char}k(P)=0$, by Hensel's lemma, $r_P/n$ is an isomorphism and the composition
\[\begin{tikzcd}
k(P)^\times / n\ar[rr, "(r_P/n)^{-1}"] && U_P/n \ar[rr, hook] && K_P^\times/n
\end{tikzcd}\]
coincides with the inflation $\operatorname{inf}: \mathrm{H}^1(k(P), \mu_n) \to \mathrm{H}^1(K_P,\mu_n)$. The same holds for $P'$ due to Lemma \ref{3.2}. Therefore, the assertion follows from the commutativity of the following diagram:
\[ \begin{tikzcd}
k(P')^\times/n \ar[r, "\cong"] \ar[d, "\varphi_P/n"'] & \mathrm{H}^1(k(P'),\mu_n) \ar[r, "\mathrm{inf}"] \ar[d]& \mathrm{H}^1(K'_{P'}, \mu_n) \ar[d, "\varphi_n"] \\
k(P)^\times/n \ar[r, "\cong"'] & \mathrm{H}^1(k(P),\mu_n)  \ar[r, "\mathrm{inf}"'] & \mathrm{H}^1(K_{P}, \mu_n)
\end{tikzcd}  \qedhere\]
\end{enumerate}
\end{proof}
